---
title: "3UTR"
author: "George Boateng-Sarfo"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(tidyr)
library(ggvenn)
library(ggpubr)
library(ggdist)
```


We used LABRAT to assess alternative 3’ UTR usage in dm6 RNA-seq data. LABRAT quantifies APA events by comparing the expression of the final two exons of every expressed transcript and calculating a ψ value that relates relative APA site usage (https://github.com/TaliaferroLab/LABRAT/). To simplify LABRAT output, only gene models with two APA isoforms were considered. For these genes, ψ values of 0 indicate exclusive usage of the proximal APA site while ψ values of 1 indicate exclusive usage of the downstream APA site.

To define tandem UTR and ALE gene models, LABRAT observes the isoform structures at the 3’ end of a gene. If all APA sites are contained within the same exon, then the structure in tandem UTR. If all APA sites are contained within different exons, then the structure is ALE. If a gene has more than two APA sites, it is possible for the gene to fit into neither classification. In these cases, LABRAT assigns the gene to have a “mixed” structure. These genes were not considered in following analyses.

To define genes that change their APA usage through neuronal differentiation by using more downstream APA sites, we utilized standard k-means clustering. Genes with changing 3’ end usage across timepoints were clustered by their mean ψ value revealing a distinct cluster of genes with increasing ψ. These genes were termed as having “neuronally restricted” (NR) long isoforms. These include 706 genes with shifts towards longer 3’ UTRs within terminal 3’ UTRs (TUTR), 269 genes with shifts towards distal alternative last exons (ALE), and 15 genes with mixed TUTR/ALE profiles; the small latter class was not considered further in our analysis. A control group of genes was defined as being expressed in the later differentiation time points (Day 47, 63, and 77) but either had no change in ψ or showed decreasing ψ values across the differentiation time course.


```{r}
# Load necessary libraries


# --- FILE PATHS ---
#labrat_file <- "~/Dropbox/3UTR/ZI_numberofposfactors.txt"
#gff3_file <- "/media/george/Extreme SSD/3UTR_RNA_Seq/LABRAT-master/annot/Drosophila_melanogaster.BDGP6.88.chr.gff3"


# --- 1. Read LABRAT output with transcript IDs ---
#labrat <- read_tsv(labrat_file, col_types = cols())

# Extract transcript IDs from 'txids' column
#labrat_txids <- labrat %>%
#  pull(txids) %>%
#  str_split("_") %>%        # split at underscores
#  unlist() %>%
#  str_trim() %>%
#  unique()

#cat("Total unique LABRAT transcripts:", length(labrat_txids), "\n")

# --- 2. Parse GFF3 annotation file ---

# Read non-header lines
gff3_lines <- readLines(gff3_file)
gff3_data <- gff3_lines[!grepl("^#", gff3_lines)]
gff3_split <- str_split_fixed(gff3_data, "\t", 9)
colnames(gff3_split) <- c("seqid", "source", "type", "start", "end", "score", "strand", "phase", "attributes")
gff3_df <- as_tibble(gff3_split)

# --- 3. Get all GFF3 transcript IDs ---
all_gff3_transcripts <- gff3_df %>%
  dplyr::filter(type == "transcript") %>%
  mutate(transcript_id = str_extract(attributes, "ID=transcript:FBtr[0-9]+"),
         transcript_id = str_remove(transcript_id, "ID=transcript:")) %>%
  dplyr::filter(!is.na(transcript_id)) %>%
  pull(transcript_id) %>%
  unique()

cat("Total transcripts in GFF3:", length(all_gff3_transcripts), "\n")

# --- 4. Get transcripts with 3′UTR annotations ---
utr3_transcripts <- gff3_df %>%
  dplyr::filter(type == "three_prime_UTR") %>%
  mutate(transcript_id = str_extract(attributes, "Parent=transcript:FBtr[0-9]+"),
         transcript_id = str_remove(transcript_id, "Parent=transcript:")) %>%
  dplyr::filter(!is.na(transcript_id)) %>%
  pull(transcript_id) %>%
  unique()

cat("Transcripts with 3'UTR annotations:", length(utr3_transcripts), "\n")

# --- 5. Overlap analyses ---
labrat_in_gff3 <- intersect(labrat_txids, all_gff3_transcripts)
labrat_not_in_gff3 <- setdiff(labrat_txids, all_gff3_transcripts)

labrat_with_3utr <- intersect(labrat_txids, utr3_transcripts)
labrat_without_3utr <- setdiff(labrat_txids, utr3_transcripts)

gff3_not_in_labrat <- setdiff(all_gff3_transcripts, labrat_txids)



# --- 8. Characterize transcripts without 3′UTRs ---

# Filter GFF3 entries that correspond to these transcripts
transcript_feature_map <- gff3_df %>%
  dplyr::filter(str_detect(attributes, "Parent=transcript:")) %>%
  mutate(transcript_id = str_extract(attributes, "Parent=transcript:FBtr[0-9]+"),
         transcript_id = str_remove(transcript_id, "Parent=transcript:")) %>%
  dplyr::filter(transcript_id %in% labrat_without_3utr)

# Summarize feature types per transcript
transcript_feature_summary <- transcript_feature_map %>%
  group_by(transcript_id) %>%
  summarise(features = paste(unique(type), collapse = ", ")) %>%
  ungroup()

# Classify based on presence/absence of specific types
transcript_feature_summary <- transcript_feature_summary %>%
  mutate(classification = case_when(
    str_detect(features, "CDS") & str_detect(features, "five_prime_UTR") ~ "CDS + 5'UTR (missing 3'UTR)",
    str_detect(features, "CDS") ~ "CDS only",
    str_detect(features, "five_prime_UTR") ~ "5'UTR only",
    str_detect(features, "exon") ~ "Exons only (noncoding?)",
    TRUE ~ "Other / Unclassified"))

# --- 6. Report results ---
cat("\n=== COMPARISON SUMMARY ===\n")
cat("LABRAT transcripts in GFF3:", length(labrat_in_gff3), "\n")
cat("LABRAT transcripts NOT in GFF3:", length(labrat_not_in_gff3), "\n\n")

cat("LABRAT transcripts with 3′UTRs:", length(labrat_with_3utr), "\n")
cat("LABRAT transcripts WITHOUT 3′UTRs:", length(labrat_without_3utr), "\n\n")

cat("GFF3 transcripts NOT found in LABRAT file:", length(gff3_not_in_labrat), "\n")


# Print summary table
table(transcript_feature_summary$classification)
# --- 7. Optional: write output files ---
#writeLines(labrat_not_in_gff3, "labrat_transcripts_missing_from_gff3.txt")
#writeLines(labrat_with_3utr, "labrat_transcripts_with_3utr.txt")
#writeLines(labrat_without_3utr, "labrat_transcripts_without_3utr.txt")
#writeLines(gff3_not_in_labrat, "gff3_transcripts_missing_from_labrat.txt")

```

```{r}
library(tximport)
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(TxDb.Dmelanogaster.UCSC.dm6.ensGene)
library(AnnotationDbi)
library(ggrepel)
# Step 1: Create tx2gene mapping from Drosophila TxDb
txdb <- TxDb.Dmelanogaster.UCSC.dm6.ensGene
k <- keys(txdb, keytype = "TXNAME")
tx2gene <- select(txdb, keys = k, columns = "GENEID", keytype = "TXNAME")

# Step 2: Define path to Salmon output directory
salmon_dir <- "/media/george/Extreme SSD/3UTR_RNA_Seq/fastq/3Seq_Salmon_outputfiles"

# Step 3: Get list of sample directories and build full paths to quant.sf files
samples <- list.dirs(path = salmon_dir, full.names = FALSE, recursive = FALSE)
files <- file.path(salmon_dir, samples, "quant.sf")
names(files) <- samples

# Optional: Check for missing files
missing <- files[!file.exists(files)]
if (length(missing) > 0) {
  stop("Missing quant.sf files:\n", paste(missing, collapse = "\n"))
}

# Step 4: Import transcript-level quantifications, summarized to gene level
txi <- tximport(files, type = "salmon", tx2gene = tx2gene)

# Step 5: Create sample metadata (adjust this if you have a metadata table)
sampleTable <- data.frame(
  row.names = colnames(txi$counts),
  condition = gsub("[0-9]+$", "", colnames(txi$counts))  # e.g., "FR109ER1" → "FR109ER"
)

# Step 6: Build DESeq2 object and normalize
dds <- DESeqDataSetFromTximport(txi, colData = sampleTable, design = ~ condition)
dds <- estimateSizeFactors(dds)

# Step 7: Variance-stabilizing transformation
vsd <- vst(dds, blind = TRUE)

# Step 8: PCA plot
pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$sample <- rownames(pcaData)

custom_colors <- c(
  "FR109ER"   = "#E41A1C",
  "FR109NR"   = "#377EB8",
  "FR112NER"  = "#4DAF4A",
  "FR112NNR"  = "#984EA3",
  "FR113NER"  = "#FF7F00",
  "FR113NNR"  = "#A65628",
  "ZI274NER"  = "#F781BF",
  "ZI274NNR"  = "#999999",
  "ZI31NER"   = "#66C2A5",
  "ZI31NNR"   = "#FC8D62",
  "ZI418NER"  = "#8DA0CB",
  "ZI418NNR"  = "#E78AC3")
ggplot(pcaData, aes(PC1, PC2, color = condition, label = sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 3, max.overlaps = 100) +
  scale_color_manual(values = custom_colors) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  ggtitle("PCA of 3'seq datasets") +  theme_classic()



library(stringr)

# Extract treatment: "ER" or "NR"
sampleTable$treatment <- sub(".*(ER|NR)[0-9]?$", "\\1", rownames(sampleTable))

# Extract strain by removing treatment and replicate info
sampleTable$strain <- sub("(ER|NR)[0-9]*$", "", rownames(sampleTable))


dds <- DESeqDataSetFromTximport(txi, colData = sampleTable, design = ~ strain + treatment)
dds <- estimateSizeFactors(dds)
vsd <- vst(dds, blind = FALSE)

library(ggrepel)

pcaData <- plotPCA(vsd, intgroup = c("strain", "treatment"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$sample <- rownames(pcaData)

ggplot(pcaData, aes(PC1, PC2, color = treatment, shape = strain, label = sample)) +
  geom_point(size = 3) +
  geom_text_repel(size = 2.5, max.overlaps = 100) +
  scale_color_manual(values = c(ER = "#E41A1C", NR = "#377EB8")) +
  theme_minimal() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA Treatment, Shaped by Strain") + theme_classic()


```


```{r}
FR109_LABRAT = read.delim("~/Dropbox/3UTR/Labrat_out/LABRAT_FR109NR2_FR109ER2.psis.pval", header = T, sep = "\t")
FR112_LABRAT = read.delim("~/Dropbox/3UTR/Labrat_out/LABRAT_FR112NNR12_FR112NER12.psis.pval", header = T, sep = "\t")
FR113_LABRAT = read.delim("~/Dropbox/3UTR/Labrat_out/LABRAT_FR113NNR2_FR113NER3.psis.pval", header = T, sep = "\t")

ZI274_LABRAT = read.delim("~/Dropbox/3UTR/Labrat_out/LABRAT_ZI274NNR1_ZI274NER2.psis.pval", header = T, sep = "\t")
ZI31_LABRAT = read.delim("~/Dropbox/3UTR/Labrat_out/LABRAT_ZI31NNR1_ZI31NER3.psis.pval", header = T, sep = "\t")
ZI418_LABRAT = read.delim("~/Dropbox/3UTR/Labrat_out/LABRAT_ZI418NNR1_ZI418NER2.psis.pval", header = T, sep = "\t")




#BC_trail = read.delim("~/Dropbox/3UTR/BC_trail.txt", header = T, sep = "\t")
#BL_trail = read.delim("~/Dropbox/3UTR/BL_trail.txt", header = T, sep = "\t")
#hiPC_human = read.delim("~/Dropbox/3UTR/hiPC_human.txt", header = T, sep = "\t")
#Seung_sup = read.delim("~/Dropbox/3UTR/Seung_Sup_hiPC", header = T, sep = "\t")
#Labrat_ZI_rep12 = read.delim("/media/george/Extreme SSD/3UTR_RNA_Seq/fastq/salmondir/LABRAT.psis.pval", header = T, sep = "\t")
#BC_trail$Gene <- sub("\\..*", "", BC_trail$Gene)
#BL_trail$Gene <- sub("\\..*", "", BL_trail$Gene)

#hiPC_human$Gene <- sub("\\..*", "", hiPC_human$Gene)


#write.table(BC_trail, file = "BC_trail.txt", quote = F, sep = "\t", col.names = T, row.names = F)
#write.table(BL_trail, file = "BL_trail.txt", quote = F, sep = "\t", col.names = T, row.names = F)


id_validation <- read.delim("~/Dropbox/3UTR/id_validation_table_35711.txt", header = TRUE, sep = "\t")
#mouse_ref = read.delim("~/Dropbox/3UTR/MGIBatchReport_20250505_164951.txt", header = TRUE, sep = "\t")

#library(biomaRt)


#ensembl_ids <- unique(hiPC_human$Gene)

#ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

#gene_map <- getBM(
#  attributes = c("ensembl_gene_id", "hgnc_symbol"),
#  filters = "ensembl_gene_id",
#  values = ensembl_ids,
#  mart = ensembl)

#BC_trail_mapped <- merge(hiPC_human, gene_map, by.x = "Gene", by.y = "ensembl_gene_id", all.x = TRUE)
#write.table(BC_trail_mapped, file = "~/Dropbox/3UTR/hiPC_human.txt", quote = F, sep = "\t", col.names = T, row.names = F)
# View a sample
#head(BC_trail_mapped)
```



```{r}
#hiPC_human <- hiPC_human %>% rename(Gene = Gene, deltapsi_hiPC = SRR10738246)
#Seung_sup   <- Seung_sup   %>% rename(Gene = Gene, deltapsi_Seung = Day02)

#merged <- inner_join(hiPC_human, Seung_sup, by = "Gene")

#merged_labeled <- merged %>%
#  mutate(psi_diff = abs(deltapsi_hiPC - deltapsi_Seung)) %>%
#  arrange(desc(psi_diff)) %>%
#  mutate(label = ifelse(row_number() <= 10, Gene, NA))


#cor_pearson <- cor.test(merged$deltapsi_hiPC, merged$deltapsi_Seung, method = "pearson")
#cor_spearman <- cor.test(merged$deltapsi_hiPC, merged$deltapsi_Seung, method = "spearman")

#r_val <- round(cor_pearson$estimate, 3)
#p_val <- signif(cor_pearson$p.value, 3)
#subtitle_text <- paste0("Pearson r = ", r_val, ", p = ", p_val)


#ggplot(merged_labeled, aes(x = deltapsi_hiPC, y = deltapsi_Seung)) +
#  geom_point(alpha = 0.65, color = "#636363", size = 2) +
#  geom_text_repel(aes(label = label), size = 3.5,box.padding = 0.4,max.overlaps = 15,segment.color = "gray40") +
#  geom_smooth(method = "lm", se = FALSE, color = "#1f78b4", linetype = "dashed", size = 0.8) +
#  geom_abline(slope = 1, intercept = 0, color = "gray60", linetype = "dotted") +
#  labs(title = expression("ΔPSI Comparison: hiPC_human vs Seung_sup (Day 2)"),
#    subtitle = subtitle_text,
#    x = expression(Delta*PSI ~ "(hiPC_human, Day 2)"),
#    y = expression(Delta*PSI ~ "(Seung_sup, Day 2)")) +
#  theme_minimal(base_size = 13) +
#  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14), plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray30"), panel.border = element_rect(color = "gray80", fill = NA),
#    aspect.ratio = 1)

```


```{r}
#hiPC_human <- hiPC_human %>%
#  rename(Gene = Gene, deltapsi_hiPC = SRR10738246)

#Seung_sup <- Seung_sup %>%
#  rename(Gene = Gene, deltapsi_Seung = Day02)


#merged <- inner_join(
#  hiPC_human %>% dplyr::select(Gene, deltapsi_hiPC, FDR, hgnc_symbol),
#  Seung_sup   %>% dplyr::select(Gene, deltapsi_Seung, cuurent_symbol),
#  by = "Gene")


#merged <- merged %>%
#  mutate(gene_label = ifelse(hgnc_symbol != "" & !is.na(hgnc_symbol), hgnc_symbol, cuurent_symbol))


#merged_labeled <- merged %>%
#  mutate(psi_diff = abs(deltapsi_hiPC - deltapsi_Seung)) %>%
#  arrange(desc(psi_diff)) %>%
#  mutate(label = ifelse(row_number() <= 10, gene_label, NA))


#cor_pearson <- cor.test(merged$deltapsi_hiPC, merged$deltapsi_Seung, method = "pearson")
#r_val <- round(cor_pearson$estimate, 3)
#p_val <- signif(cor_pearson$p.value, 3)
#subtitle_text <- paste0("Pearson r = ", r_val, ", p = ", p_val)


#ggplot(merged_labeled, aes(x = deltapsi_hiPC, y = deltapsi_Seung)) +
#  geom_point(alpha = 0.65, color = "#636363", size = 2) +
#  geom_text_repel(aes(label = label), size = 3.5,
#                  box.padding = 0.4, max.overlaps = 15, segment.color = "gray40") +
#  geom_smooth(method = "lm", se = FALSE, color = "#1f78b4", linetype = "dashed", size = 0.8) +
#  geom_abline(slope = 1, intercept = 0, color = "gray60", linetype = "dotted") +
#  labs(title = expression("ΔPSI Comparison: hiPC_human vs Seung_sup (Day 2)"),
#    subtitle = subtitle_text,
#    x = expression(Delta*PSI ~ "(hiPC_human, Day 2)"),
#    y = expression(Delta*PSI ~ "(Seung_sup, Day 2)")) +
#  theme_minimal(base_size = 13) +
#  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
#    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray30"),
#    panel.border = element_rect(color = "gray80", fill = NA),
#    aspect.ratio = 1)

```



```{r}
# Count significant genes in hiPC_human
#n_sig_hipc <- hiPC_human %>%
#  dplyr::filter(FDR < 0.05) %>%
#  nrow()

# Count genes that exist in both datasets with valid ΔPSI
#n_sig_in_both <- merged %>%
#  dplyr::filter(FDR < 0.05 & !is.na(deltapsi_hiPC) & !is.na(deltapsi_Seung)) %>%
#  nrow()

# Get top 50 as before
#sig_genes <- merged %>%
#  dplyr::filter(FDR < 0.05 & !is.na(deltapsi_hiPC) & !is.na(deltapsi_Seung)) %>%
#  mutate(abs_dpsi = abs(deltapsi_hiPC)) %>%
#  arrange(desc(abs_dpsi)) %>%
#  slice_head(n = 50)

# Prepare long format
#sig_long <- sig_genes %>%
#  dplyr::select(gene_label, deltapsi_hiPC, deltapsi_Seung) %>%
#  pivot_longer(cols = c(deltapsi_hiPC, deltapsi_Seung),
#               names_to = "Dataset", values_to = "Delta_PSI") %>%
#  mutate(Dataset = recode(Dataset,
#                          "deltapsi_hiPC" = "hiPC_human",
#                          "deltapsi_Seung" = "Seung_sup"))

# Order genes by hiPC_human ΔPSI
#gene_order <- sig_long %>%
#  dplyr::filter(Dataset == "hiPC_human") %>%
#  arrange(Delta_PSI) %>%
 # pull(gene_label)
#sig_long$gene_label <- factor(sig_long$gene_label, levels = gene_order)

# Build subtitle text
#subtitle_text <- paste0(
#  "Significant in hiPC_human (FDR < 0.05): ", n_sig_hipc, 
#  " genes | Present in both datasets: ", n_sig_in_both)

# Plot
#ggplot(sig_long, aes(x = gene_label, y = Delta_PSI, fill = Dataset)) +
#  geom_bar(stat = "identity", position = "dodge") +
#  coord_flip() +
#  labs(
#    title = "Top 50 Genes: ΔPSI in hiPC_human and Seung_sup (Day 2)",
#    subtitle = subtitle_text,
#    x = "Gene Symbol",
#    y = expression(Delta*PSI),
#    fill = "Dataset") +
#  scale_fill_manual(values = c("hiPC_human" = "#E41A1C", "Seung_sup" = "#377EB8")) +
#  theme_minimal(base_size = 13) +
#  theme(
#    plot.title = element_text(face = "bold", hjust = 0.5),
#    plot.subtitle = element_text(hjust = 0.5, size = 11, color = "gray40"),
#    legend.position = "top",
#    panel.grid.minor = element_blank())


```



```{r}

# Prepare data
#psi_violin_data <- merged %>%
#  dplyr::select(gene_label, psi_hiPC = deltapsi_hiPC, psi_Seung = deltapsi_Seung) %>%
#  pivot_longer(cols = c(psi_hiPC, psi_Seung),
#               names_to = "Dataset", values_to = "PSI") %>%
#  mutate(Dataset = recode(Dataset,
#                          "psi_hiPC" = "hiPC_human (SRR10738246)",
#                          "psi_Seung" = "Seung_sup (Day02)"))

# Violin plot with boxplot overlay and statistical annotation
#ggplot(psi_violin_data, aes(x = Dataset, y = PSI, fill = Dataset)) +
#  geom_violin(trim = FALSE, alpha = 0.6, color = NA) +
#  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8, notch = T) +
#  stat_compare_means(method = "wilcox.test", label.y = 1.05, size = 4, label.x = 1.1) +
#  labs(
#    title = "PSI Distribution in hiPC_human vs Seung_sup (Day 2)",
#    x = "",
#    y = expression(Delta*PSI)) +
#  scale_fill_manual(values = c(
#    "hiPC_human (SRR10738246)" = "#E41A1C",
#    "Seung_sup (Day02)" = "#377EB8")) +
#  theme_minimal(base_size = 13) +
#  theme(
#    plot.title = element_text(hjust = 0.5, face = "bold"),
#    legend.position = "none",
#    panel.grid.minor = element_blank())
```



```{r}

FR109_LABRAT = left_join(FR109_LABRAT, id_validation, by = c("Gene" = "submitted_item"))
FR112_LABRAT = left_join(FR112_LABRAT, id_validation, by = c("Gene" = "submitted_item"))
FR113_LABRAT = left_join(FR113_LABRAT, id_validation, by =c ("Gene" = "submitted_item"))

ZI274_LABRAT = left_join(ZI274_LABRAT, id_validation, by = c("Gene" = "submitted_item"))
ZI31_LABRAT = left_join(ZI31_LABRAT, id_validation, by = c("Gene" = "submitted_item"))
ZI418_LABRAT = left_join(ZI418_LABRAT, id_validation, by =c ("Gene" = "submitted_item"))

FR109_LABRAT = FR109_LABRAT %>%
  distinct(validated_id, .keep_all = T)

FR112_LABRAT = FR112_LABRAT %>%
  distinct(validated_id, .keep_all = T)

FR113_LABRAT = FR113_LABRAT %>%
  distinct(validated_id, .keep_all = T)

ZI274_LABRAT = ZI274_LABRAT %>%
  distinct(validated_id, .keep_all = T)

ZI31_LABRAT = ZI31_LABRAT %>%
  distinct(validated_id, .keep_all = T)

ZI418_LABRAT = ZI418_LABRAT %>%
  distinct(validated_id, .keep_all = T)





#BL_trail <- left_join(BL_trail, mouse_ref, by = c("Gene" = "Input"))
```


The interpretation of Δψ (deltapsi = Condition B - Condition A) in the LABRAT analysis is based on how the structure of the input. 

How LABRAT Calculates Δψ
Input Requirements:

The sample metadata file (--samples) includes:

A column condition with two groups (e.g., FR1 and Z1).

Explicit flags --conditionA and --conditionB when running LABRAT (e.g., --conditionA FR1 --conditionB Z1).

Δψ Calculation:

LABRAT computes:

Δψ = ψ FR1 (French) - ψZ1 (Zambian) 
 
This is always Condition B - Condition A, as specified in the command.

Interpreting Δψ in the Context of our analysis:

Δψ Value	Proximal/Distal usage	of 3'UTR in the Zambian vs. French
Δψ > 0	Higher proximal usage in Z1 3'UTRs in Zambian population (relative to French)
Δψ < 0	Higher distal usage in Z1 3'UTRs in Zambian population (relative to French)
Δψ ≈ 0	No difference	No significant APA change between populations


Directionality Consistency:

All your results (volcano plots, tables, etc.) follow this convention:

Positive Δψ → Z1 > FR1 (Zambian prefers proximal sites).

Negative Δψ → Z1 < FR1 (Zambian prefers distal sites).

Biological Implications:

If you see Δψ = +0.2 for a gene:

The Zambian population has 20% higher proximal site usage than the French.

This gene’s mRNA may escape miRNA repression in Zambians due to shorter 3'UTRs.

If Δψ = -0.3:
```{r}


#FR109_LABRAT,FR112_LABRAT,FR113_LABRAT,ZI274_LABRAT,ZI31_LABRAT,ZI418_LABRAT



FR109_filtered <- FR109_LABRAT %>%
  dplyr::filter(FDR <= 0.05)

FR112_filtered <- FR112_LABRAT %>%
  dplyr::filter(FDR <= 0.05)

FR113_filtered <- FR113_LABRAT %>%
  dplyr::filter(FDR <= 0.05)

ZI274_filtered <- ZI274_LABRAT %>%
  dplyr::filter(FDR <= 0.05)

ZI31_filtered <- ZI31_LABRAT %>%
  dplyr::filter(FDR <= 0.05)

ZI418_filtered <- ZI418_LABRAT %>%
  dplyr::filter(FDR <= 0.05)

```

```{r}
all_sig_genes <- bind_rows(
  FR109_filtered,
  FR112_filtered,
  FR113_filtered,
  ZI274_filtered,
  ZI31_filtered,
  ZI418_filtered)


length(unique(all_sig_genes$validated_id))


# Create French and Zambian sets
fr_ids <- unique(c(
  FR109_filtered$validated_id,
  FR112_filtered$validated_id,
  FR113_filtered$validated_id))

za_ids <- unique(c(
  ZI274_filtered$validated_id,
  ZI31_filtered$validated_id,
  ZI418_filtered$validated_id))

# Totals
length(fr_ids)
length(za_ids)

# Shared
length(intersect(fr_ids, za_ids))

# Unique to each
length(setdiff(fr_ids, za_ids))       # French-only
length(setdiff(za_ids, fr_ids))       # Zambian-only


# Check genetype composition among all significant genes
all_sig_genes %>%
  distinct(validated_id, .keep_all = TRUE) %>%
  dplyr::count(genetype)


# Assign ChangeType to each gene
all_sig_genes <- all_sig_genes %>%
  mutate(ChangeType = case_when(
    validated_id %in% fr_ids & validated_id %in% za_ids ~ "Shared",
    validated_id %in% fr_ids ~ "French-only",
    validated_id %in% za_ids ~ "Zambian-only",
    TRUE ~ "Unknown"))

# Summarize by APA type and ChangeType
all_sig_genes %>%
  distinct(validated_id, .keep_all = TRUE) %>%
  dplyr::count(ChangeType, genetype)

```


```{r}
FR_filtered <- bind_rows(FR109_filtered, FR112_filtered, FR113_filtered)
ZA_filtered <- bind_rows(ZI274_filtered, ZI31_filtered, ZI418_filtered)

FR_pref <- FR_filtered %>%
  mutate(APA_preference = ifelse(deltapsi > 0, "Distal", "Proximal"))

ZA_pref <- ZA_filtered %>%
  mutate(APA_preference = ifelse(deltapsi > 0, "Distal", "Proximal"))

FR_pref %>% dplyr::count(APA_preference)
ZA_pref %>% dplyr::count(APA_preference)


FR_pref_dedup <- FR_pref %>%
  distinct(validated_id, .keep_all = TRUE) %>%
  dplyr::count(APA_preference)

ZA_pref_dedup <- ZA_pref %>%
  distinct(validated_id, .keep_all = TRUE) %>%
  dplyr::count(APA_preference)



```


```{r}


FR_all <- bind_rows(FR109_filtered, FR112_filtered, FR113_filtered)
FR_collapsed <- FR_all %>%
  group_by(Gene) %>%
  summarise(FRER1 = mean(c_across(contains("NER")), na.rm = TRUE), FRNR1 = mean(c_across(contains("NNR")), na.rm = TRUE),
    deltapsi = mean(deltapsi, na.rm = TRUE), pval = mean(pval, na.rm = TRUE),
    FDR = mean(FDR, na.rm = TRUE), validated_id = first(validated_id),
    current_symbol = first(current_symbol),
    genetype = first(genetype)) %>%
  ungroup()


ZA_all <- bind_rows(ZI274_filtered, ZI31_filtered, ZI418_filtered)

ZA_collapsed <- ZA_all %>%
  group_by(Gene) %>%
  summarise(ZAER1 = mean(c_across(contains("NER")), na.rm = TRUE), ZANR1 = mean(c_across(contains("NNR")), na.rm = TRUE),
    deltapsi = mean(deltapsi, na.rm = TRUE), pval = mean(pval, na.rm = TRUE),
    FDR = mean(FDR, na.rm = TRUE), validated_id = first(validated_id),
    current_symbol = first(current_symbol),
    genetype = first(genetype)) %>%
  ungroup()


FR_collapsed <- FR_collapsed %>% 
  distinct(validated_id, .keep_all = TRUE)
ZA_collapsed <- ZA_collapsed %>% 
  distinct(validated_id, .keep_all = TRUE)

#write.table(ZA_collapsed, file = "Supp_ZA_collapsed.txt", sep = "\t", quote = F, row.names = F, col.names = T)

```

Positive Δψ (> 0):
The control (NR) group has a higher ψ value than the treatment (ER)

More usage of distal poly(A) sites in NR vs ER

Treatment (ER) causes a shift toward proximal APA sites (shorter 3′ UTRs)

egative Δψ (< 0):
The treatment (ER) group has a higher ψ value than the control (NR)

More usage of distal poly(A) sites in ER vs NR

Treatment promotes lengthening of 3′ UTRs (more distal APA)

Since Δψ = ψ(NR) - ψ(ER), a positive value means treatment shortens the 3′ UTR, and a negative value means treatment lengthens the 3′ UTR.
```{r}

FR_bar_data <- FR_collapsed %>%
  mutate(genetype = as.character(genetype)) %>%
  dplyr::count(genetype) %>%
  dplyr::rename(Count = n)

ZA_bar_data <- ZA_collapsed %>%
  mutate(genetype = as.character(genetype)) %>%
  dplyr::count(genetype) %>%
  dplyr::rename(Count = n)


ggplot(FR_bar_data, aes(x = genetype, y = Count, fill = genetype)) +
  geom_bar(stat = "identity", width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = Count), vjust = -0.5, size = 5) +
  labs(title = "APA Genetype Distribution (French)",
       x = "Genetype", y = "Number of Genes") +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  theme_classic(base_size = 14) +   
  scale_fill_brewer(palette = "Set2")


ggplot(ZA_bar_data, aes(x = genetype, y = Count, fill = genetype)) +
  geom_bar(stat = "identity", width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = Count), vjust = -0.5, size = 5) +
  labs(title = "APA Genetype Distribution (Zambian)",
       x = "Genetype", y = "Number of Genes") +
  theme_minimal(base_size = 14) +
  theme_classic(base_size = 14) +   
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  scale_fill_brewer(palette = "Set2")


FR_bar_data <- FR_collapsed %>%
  mutate(genetype = as.character(genetype)) %>%
  dplyr::count(genetype) %>%
  dplyr::rename(Count = n) %>%
  mutate(Population = "French")

ZA_bar_data <- ZA_collapsed %>%
  mutate(genetype = as.character(genetype)) %>%
  dplyr::count(genetype) %>%
  dplyr::rename(Count = n) %>%
  mutate(Population = "Zambian")


combined_bar_data <- bind_rows(FR_bar_data, ZA_bar_data)

ggplot(combined_bar_data, aes(x = genetype, y = Count, fill = Population)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.6), width = 0.5) +
  geom_text(aes(label = Count), 
            position = position_dodge(width = 0.6), 
            vjust = -0.5, size = 5) +
  labs(title = "APA genetype distribution by Population",
    x = "Genetype", y = "# of genes") +
  theme_classic(base_size = 14) +
  theme(panel.grid = element_blank(),
    axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  scale_fill_brewer(palette = "Set1")
```


```{r}
# Create counts per genotype (not population)
FR109_bar <- FR109_filtered %>%
  mutate(genetype = as.character(genetype)) %>%
  count(genetype) %>%
  rename(Count = n) %>%
  mutate(Genotype = "FR109")

FR112_bar <- FR112_filtered %>%
  mutate(genetype = as.character(genetype)) %>%
  count(genetype) %>%
  rename(Count = n) %>%
  mutate(Genotype = "FR112")

FR113_bar <- FR113_filtered %>%
  mutate(genetype = as.character(genetype)) %>%
  count(genetype) %>%
  rename(Count = n) %>%
  mutate(Genotype = "FR113")

ZI274_bar <- ZI274_filtered %>%
  mutate(genetype = as.character(genetype)) %>%
  count(genetype) %>%
  rename(Count = n) %>%
  mutate(Genotype = "ZI274")

ZI31_bar <- ZI31_filtered %>%
  mutate(genetype = as.character(genetype)) %>%
  count(genetype) %>%
  rename(Count = n) %>%
  mutate(Genotype = "ZI31")

ZI418_bar <- ZI418_filtered %>%
  mutate(genetype = as.character(genetype)) %>%
  count(genetype) %>%
  rename(Count = n) %>%
  mutate(Genotype = "ZI418")

# Combine all into one data frame
all_genotype_bar_data <- bind_rows(
  FR109_bar, FR112_bar, FR113_bar,
  ZI274_bar, ZI31_bar, ZI418_bar)
ggplot(all_genotype_bar_data, aes(x = Genotype, y = Count, fill = genetype)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = Count), position = position_stack(vjust = 0.5), size = 4, color = "white") +
  labs(title = "APA Genetype Distribution by Genotype",
       x = "Genotype", y = "# of Genes", fill = "Genetype") +
  theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  scale_fill_brewer(palette = "Set2")

```


```{r}
FR_box <- FR_collapsed %>%
  mutate(Population = "French",
         Direction = ifelse(deltapsi >= 0, "Distal APA usage", "Proximal APA usage"))

ZA_box <- ZA_collapsed %>%
  mutate(Population = "Zambian",
         Direction = ifelse(deltapsi >= 0, "Distal APA usage", "Proximal APA usage"))


box_data <- bind_rows(FR_box, ZA_box)


box_data$Direction <- factor(box_data$Direction, levels = c("Distal APA usage", "Proximal APA usage"))
box_data$Population <- factor(box_data$Population, levels = c("French", "Zambian"))

# Split data
neg_data <- box_data %>% 
  dplyr::filter(Direction == "Proximal APA usage")
pos_data <- box_data %>% 
  dplyr::filter(Direction == "Distal APA usage")

# Plot for Negative Δpsi FR_volcano <- FR_volcano %>%  dplyr::filter(current_symbol != "UQCR-Q")
ggplot(neg_data,aes(x = Population, y = deltapsi, fill = Population)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = TRUE) +
  geom_jitter(aes(color = Population), width = 0.20, alpha = 0.7) +
  stat_compare_means(
  comparisons = list(c("French", "Zambian")),
  method = "wilcox.test", label = "p.signif", label.y = 0.1) +
  labs(title = " ", y     = "ΔΨ (relative to control samples)", x     = "") +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 14) +          # keeps axis lines, no grid
  theme(legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)) + coord_flip()


ggplot(pos_data, aes(x = Population, y = deltapsi, fill = Population)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = T) +
  geom_jitter(aes(color = Population), width = 0.2, alpha = 0.7) + 
  stat_compare_means(
  comparisons = list(c("French", "Zambian")),
  method = "wilcox.test", label = "p.signif", label.y = 0.1) +
  labs(title = " ", y     = "ΔΨ (relative to control samples)", x     = "") +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 14) +          # keeps axis lines, no grid
  theme(legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)) + coord_flip()

# Combine and plot both directions in one figure
ggplot(box_data, aes(x = Population, y = deltapsi, fill = Population)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = TRUE) +
  geom_jitter(aes(color = Population), width = 0.2, alpha = 0.7, color = "black") +
  stat_compare_means(method = "wilcox.test", aes(group = Population), label = "p.signif", label.y = 0.1,
    group.by = "Direction") +
  facet_wrap(~Direction, ncol = 1, scales = "free_y", strip.position = "right") +
  labs(title = "", y = "ΔΨ (relative to control samples)",x = "") +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  coord_flip()
```



 ΔΨ values within each population, the question changes from "Are French and Zambian APA profiles different?" to:

"Within a given population, do genes that shift toward distal sites (positive ΔΨ) differ significantly from those shifting toward proximal sites (negative ΔΨ)?"



```{r}

FR_box <- box_data %>% filter(Population == "French")
ZA_box <- box_data %>% filter(Population == "Zambian")

# French population
wilcox.test(deltapsi ~ Direction, data = FR_box)

# Zambian population
wilcox.test(deltapsi ~ Direction, data = ZA_box)

base_colors <- c("Proximal" = "pink", "Distal" = "darkblue")

ggplot(box_data, aes(x = Direction, y = deltapsi, fill = Direction)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = TRUE) +
  geom_jitter(width = 0.2, alpha = 0.7) + ylim(-0.5, 0.5) +
  stat_compare_means(method = "wilcox.test", aes(group = Direction), label = "p.signif") +
  facet_wrap(~Population, ncol = 1, strip.position = "right") +
  labs(y = "ΔΨ", x = "APA Direction", title = "Within-Population ΔΨ Differences") +
  scale_color_manual(values = base_colors)+
  theme_classic(base_size = 14) + coord_flip()

```


```{r}


ZA_all <- bind_rows(ZI274_LABRAT, ZI31_LABRAT, ZI418_LABRAT)

# Collapse: keep the entry with lowest FDR per gene
ZA_collapsed <- ZA_all %>%
  group_by(validated_id) %>%
  arrange(FDR) %>%
  slice(1) %>%
  ungroup()

# Define FDR threshold
fdr_thresh <- 0.05

# Volcano plot input
ZA_volcano <- ZA_collapsed %>%
  mutate(sig_cat = case_when(FDR < fdr_thresh & deltapsi > 0 ~ "Proximal", FDR < fdr_thresh & deltapsi < 0 ~ "Distal", TRUE ~ "Not significant"), log10_pval = -log10(FDR)) %>%
  mutate(log10_pval = ifelse(is.infinite(log10_pval), NA, log10_pval))

# Count categories for legend
ZA_counts <- ZA_volcano %>%
  count(sig_cat) %>%
  mutate(label = paste0(sig_cat, " (n=", n, ")"))

legend_labels <- setNames(ZA_counts$label, ZA_counts$sig_cat)

# Label top 10 genes
ZA_labeled <- ZA_volcano %>%
  filter(FDR < fdr_thresh) %>%
  slice_max(abs(deltapsi), n = 10)


base_colors <- c("Proximal" = "pink", "Distal" = "darkblue", "Not significant" = "gray70")


ggplot(ZA_volcano, aes(x = deltapsi, y = log10_pval, color = sig_cat)) +
  geom_point(alpha = 0.8, size = 3) + 
  ylim(0,5) +
  geom_text_repel(data = ZA_labeled, aes(label = current_symbol, colour = "black"), size = 3, box.padding = 0.4) +
  scale_color_manual(values = base_colors, labels = legend_labels) +
  geom_hline(yintercept = -log10(fdr_thresh), linetype = "dashed", color = "black") +
  labs(title = "Zambian Population: APA Changes", x = expression(Delta*psi~"(ψ[NR] - ψ[ER])"), y = expression(-log[10](FDR)), color = "APA Category") +
  theme_classic(base_size = 14) +
  theme(legend.position = "right", plot.title = element_text(face = "bold", hjust = 0.5))

```



```{r}
# Combine all raw replicate data
FR_all <- bind_rows(FR109_LABRAT, FR112_LABRAT, FR113_LABRAT)

# Collapse: keep the entry with lowest FDR per gene
FR_collapsed <- FR_all %>%
  group_by(validated_id) %>%
  arrange(FDR) %>%
  slice(1) %>%
  ungroup()

# Define FDR threshold
fdr_thresh <- 0.05

# Volcano plot input
FR_volcano <- FR_collapsed %>%
  mutate(sig_cat = case_when(FDR < fdr_thresh & deltapsi > 0 ~ "Proximal", FDR < fdr_thresh & deltapsi < 0 ~ "Distal", TRUE ~ "Not significant"), log10_pval = -log10(FDR)) %>%
  mutate(log10_pval = ifelse(is.infinite(log10_pval), NA, log10_pval))

# Count categories for legend
FR_counts <- FR_volcano %>%
  count(sig_cat) %>%
  mutate(label = paste0(sig_cat, " (n=", n, ")"))

legend_labels <- setNames(FR_counts$label, FR_counts$sig_cat)

# Label top 10 genes
FR_labeled <- FR_volcano %>%
  filter(FDR < fdr_thresh) %>%
  slice_max(abs(deltapsi), n = 10)


#base_colors <- c("Proximal" = "#D55E00", "Distal" = "#0072B2", "Not significant" = "gray70")


ggplot(FR_volcano, aes(x = deltapsi, y = log10_pval, color = sig_cat)) +
  geom_point(alpha = 0.8, size = 2) + ylim(0,5) +
  geom_text_repel(data = FR_labeled, aes(label = current_symbol), size = 3, box.padding = 0.4, colour = "black") +
  scale_color_manual(values = base_colors, labels = legend_labels) +
  geom_hline(yintercept = -log10(fdr_thresh), linetype = "dashed", color = "black") +
  labs(title = "French Population: APA Changes", x = expression(Delta*psi~"(ψ[NR] - ψ[ER])"), y = expression(-log[10](FDR)), color = "APA Category") +
  theme_classic(base_size = 14) +
  theme(legend.position = "right", plot.title = element_text(face = "bold", hjust = 0.5))

```



```{r}
FR109 <- FR109_LABRAT %>% mutate(Genotype = "FR109")
FR112 <- FR112_LABRAT %>% mutate(Genotype = "FR112")
FR113 <- FR113_LABRAT %>% mutate(Genotype = "FR113")

FR_all <- bind_rows(FR109, FR112, FR113)
FR_collapsed <- FR_all %>%
  group_by(Genotype, validated_id) %>%
  arrange(FDR) %>%
  slice(1) %>%
  ungroup()

fdr_thresh <- 0.05

FR_volcano <- FR_collapsed %>%
  mutate(sig_cat = case_when(
    FDR < fdr_thresh & deltapsi > 0 ~ "Proximal",
    FDR < fdr_thresh & deltapsi < 0 ~ "Distal",
    TRUE ~ "Not significant"
  ),
  log10_pval = -log10(FDR),
  log10_pval = ifelse(is.infinite(log10_pval), NA, log10_pval))

FR_labeled <- FR_volcano %>%
  filter(FDR < fdr_thresh) %>%
  group_by(Genotype) %>%
  slice_max(abs(deltapsi), n = 3) %>%
  ungroup()

ggplot(FR_volcano, aes(x = deltapsi, y = log10_pval, color = sig_cat, shape = Genotype)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel(data = FR_labeled,
                  aes(label = current_symbol),
                  size = 3, box.padding = 0.4, max.overlaps = 100,
                  color = "black", show.legend = FALSE) +
  geom_hline(yintercept = -log10(fdr_thresh), linetype = "dashed", color = "black") +
  scale_color_manual(values = c(
    "Proximal" = "pink",
    "Distal" = "darkblue",
    "Not significant" = "gray70")) +
  scale_shape_manual(values = c(16, 17, 15)) +  # circles, triangles, squares
  labs(
    title = "APA Volcano Plot by Genotype (French)",
    x = expression(Delta*psi~"(ψ[NR] - ψ[ER])"),
    y = expression(-log[10](FDR)),
    color = "Significance",
    shape = "Genotype") +
  theme_classic(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), legend.key.size = unit(0.9, "lines"))

FR_counts <- FR_volcano %>%
  count(sig_cat) %>%
  mutate(label = paste0(sig_cat, " (n=", n, ")"))

# Named vector for legend labeling
legend_labels <- setNames(FR_counts$label, FR_counts$sig_cat)

volcano_low <- FR_volcano %>% filter(log10_pval <= 5)
volcano_high <- FR_volcano %>% filter(log10_pval >= 50)
p1 <- ggplot(volcano_low, aes(x = deltapsi, y = log10_pval, color = sig_cat, shape = Genotype)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel(data = filter(FR_labeled, log10_pval <= 5),
                  aes(label = current_symbol),
                  size = 3, box.padding = 0.4, max.overlaps = 50,
                  color = "black", show.legend = FALSE) +
  geom_hline(yintercept = -log10(fdr_thresh), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Proximal" = "pink", "Distal" = "darkblue", "Not significant" = "gray70"),
                     labels = legend_labels) +
  scale_shape_manual(values = c("FR109" = 16, "FR112" = 17, "FR113" = 15)) +
  scale_y_continuous(breaks = seq(0, 5, 1), limits = c(0, 5)) +
  labs(
    y = expression(-log[10](FDR)),
    x = expression(Delta*psi~"(ψ[NR] - ψ[ER])")) +
  theme_classic(base_size = 14) +
  theme(legend.position = "right")

p2 <- ggplot(volcano_high, aes(x = deltapsi, y = log10_pval, color = sig_cat, shape = Genotype)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel(data = filter(FR_labeled, log10_pval >= 50),
                  aes(label = current_symbol),
                  size = 3, box.padding = 0.4, max.overlaps = 50,
                  color = "black", show.legend = FALSE) +
  scale_color_manual(values = c("Proximal" = "pink", "Distal" = "darkblue", "Not significant" = "gray70"),
                     labels = legend_labels) +
  scale_shape_manual(values = c("FR109" = 16, "FR112" = 17, "FR113" = 15)) +
  coord_cartesian(ylim = c(50, 60)) +
  scale_y_continuous(breaks = c(50, 55, 60)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none",
    plot.margin = margin(b = 0)
  )


library(patchwork)

final_volcano <- p2 / p1 + 
  plot_layout(heights = c(1, 3)) +
  plot_annotation(title = "APA Volcano Plot by Genotype (French)") &
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
final_volcano
```


```{r}
# Add Genotype labels
ZI274 <- ZI274_LABRAT %>% mutate(Genotype = "ZI274")
ZI31  <- ZI31_LABRAT  %>% mutate(Genotype = "ZI31")
ZI418 <- ZI418_LABRAT %>% mutate(Genotype = "ZI418")

# Combine into one data frame
ZA_all <- bind_rows(ZI274, ZI31, ZI418)
ZA_collapsed <- ZA_all %>%
  group_by(Genotype, validated_id) %>%
  arrange(FDR) %>%
  slice(1) %>%
  ungroup()

fdr_thresh <- 0.05

ZA_volcano <- ZA_collapsed %>%
  mutate(sig_cat = case_when(
    FDR < fdr_thresh & deltapsi > 0 ~ "Proximal",
    FDR < fdr_thresh & deltapsi < 0 ~ "Distal",
    TRUE ~ "Not significant"),
  log10_pval = -log10(FDR),
  log10_pval = ifelse(is.infinite(log10_pval), NA, log10_pval))

ZA_counts <- ZA_volcano %>%
  count(sig_cat) %>%
  mutate(label = paste0(sig_cat, " (n=", n, ")"))

legend_labels <- setNames(ZA_counts$label, ZA_counts$sig_cat)

ZA_labeled <- ZA_volcano %>%
  filter(FDR < fdr_thresh) %>%
  group_by(Genotype) %>%
  slice_max(abs(deltapsi), n = 3) %>%
  ungroup()

volcano_low <- ZA_volcano %>% filter(log10_pval <= 5)
volcano_high <- ZA_volcano %>% filter(log10_pval >= 50)


p1 <- ggplot(volcano_low, aes(x = deltapsi, y = log10_pval, color = sig_cat, shape = Genotype)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel(data = filter(ZA_labeled, log10_pval <= 5),
                  aes(label = current_symbol),
                  size = 3, box.padding = 0.4, max.overlaps = 50,
                  color = "black", show.legend = FALSE) +
  geom_hline(yintercept = -log10(fdr_thresh), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Proximal" = "pink", "Distal" = "darkblue", "Not significant" = "gray70"),
                     labels = legend_labels) +
  scale_shape_manual(values = c("ZI274" = 16, "ZI31" = 17, "ZI418" = 15)) +
  scale_y_continuous(breaks = seq(0, 5, 1), limits = c(0, 5)) +
  labs(
    y = expression(-log[10](FDR)),
    x = expression(Delta*psi~"(ψ[NR] - ψ[ER])")) +
  theme_classic(base_size = 14) +
  theme(legend.position = "right")

p2 <- ggplot(volcano_high, aes(x = deltapsi, y = log10_pval, color = sig_cat, shape = Genotype)) +
  geom_point(alpha = 0.8, size = 3) +
  geom_text_repel(data = filter(ZA_labeled, log10_pval >= 50),
                  aes(label = current_symbol),
                  size = 3, box.padding = 0.4, max.overlaps = 50,
                  color = "black", show.legend = FALSE) +
  scale_color_manual(values = c("Proximal" = "pink", "Distal" = "darkblue", "Not significant" = "gray70"),
                     labels = legend_labels) +
  scale_shape_manual(values = c("ZI274" = 16, "ZI31" = 17, "ZI418" = 15)) +
  coord_cartesian(ylim = c(50, 60)) +
  scale_y_continuous(breaks = c(50, 55, 60)) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none",
    plot.margin = margin(b = 0))



final_plot_ZA <- p2 / p1 +
  plot_layout(heights = c(1, 3)) +
  plot_annotation(title = "APA Volcano Plot by Genotype (Zambian)") &
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

final_plot_ZA
```

LABRAT calculates:

Δψ = ψ_ER − ψ_NR

So:

If Δψ > 0 → Higher distal APA usage in ER (treatment) compared to NR (control)

If Δψ < 0 → Higher proximal APA usage in ER (treatment) compared to NR (control)


| Δψ sign | Interpretation                                    |
| ------- | ------------------------------------------------- |
| **> 0** | ER (treatment) favors **distal** APA site usage   |
| **< 0** | ER (treatment) favors **proximal** APA site usage |

APA Site Usage ↔ 3′UTR Length
Proximal APA site → Shorter 3′UTR

Distal APA site → Longer 3′UTR

| Δψ value | Treatment (ER) APA usage | 3′UTR length change in treatment |
| -------- | ------------------------ | -------------------------------- |
| **> 0**  | More **distal** usage    | **Lengthening** of 3′UTR         |
| **< 0**  | More **proximal** usage  | **Shortening** of 3′UTR          |


```{r}
FR_posfactors_109 <- read.delim("~/Dropbox/3UTR/Labrat_out/FR109NR2_FR109ER2_numberofposfactors.txt", header = TRUE, sep = "\t") %>%
  mutate(Genotype = "FR109")
FR_posfactors_112 <- read.delim("~/Dropbox/3UTR/Labrat_out/FR112NNR12_FR112NER12_numberofposfactors.txt", header = TRUE, sep = "\t") %>%
  mutate(Genotype = "FR112")
FR_posfactors_113 <- read.delim("~/Dropbox/3UTR/Labrat_out/FR113NNR2_FR113NER3_numberofposfactors.txt", header = TRUE, sep = "\t") %>%
  mutate(Genotype = "FR113")

ZI_posfactors_ZI274 <- read.delim("~/Dropbox/3UTR/Labrat_out/ZI274NNR1_ZI274NER2_numberofposfactors.txt", header = TRUE, sep = "\t") %>%
  mutate(Genotype = "ZI274")
ZI_posfactors_ZI31 <- read.delim("~/Dropbox/3UTR/Labrat_out/ZI31NNR1_ZI31NER3_numberofposfactors.txt", header = TRUE, sep = "\t") %>%
  mutate(Genotype = "ZI31")
ZI_posfactors_ZI418 <- read.delim("~/Dropbox/3UTR/Labrat_out/ZI418NER2_ZI418NER2_numberofposfactors.txt", header = TRUE, sep = "\t") %>%
  mutate(Genotype = "ZI418")


FR109_sig <- FR109_LABRAT %>%
  filter(FDR < 0.05) %>%
  left_join(FR_posfactors_109, by = "Gene")

FR112_sig <- FR112_LABRAT %>%
  filter(FDR < 0.05) %>%
  left_join(FR_posfactors_112, by = "Gene")

FR113_sig <- FR113_LABRAT %>%
  filter(FDR < 0.05) %>%
  left_join(FR_posfactors_113, by = "Gene")

ZI274_sig <- ZI274_LABRAT %>%
  filter(FDR < 0.05) %>%
  left_join(ZI_posfactors_ZI274, by = "Gene")

ZI31_sig <- ZI31_LABRAT %>%
  filter(FDR < 0.05) %>%
  left_join(ZI_posfactors_ZI31, by = "Gene")

ZI418_sig <- ZI418_LABRAT %>%
  filter(FDR < 0.05) %>%
  left_join(ZI_posfactors_ZI418, by = "Gene")


FR_combined <- bind_rows(FR109_sig, FR112_sig, FR113_sig)
ZA_combined <- bind_rows(ZI274_sig, ZI31_sig, ZI418_sig)


apa_sig_genes <- bind_rows(FR_combined, ZA_combined) %>%
  filter(numberofposfactors > 1)


apa_counts <- apa_sig_genes %>%
  count(Genotype, numberofposfactors) %>%
  rename(`APA Sites` = numberofposfactors, `Number of Genes` = n)


ggplot(apa_counts, aes(x = factor(`APA Sites`), y = `Number of Genes`, fill = Genotype)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = `Number of Genes`),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4) +
  labs(
    title = "APA Site Count Distribution in Significant Genes (FDR < 0.05)",
    x = "Number of APA Sites (≥25 nt apart)",
    y = "Number of Genes") +
  scale_fill_brewer(palette = "Set1") +
  theme_classic(base_size = 14)
```


Do APA shifts toward distal or proximal sites differ in their genomic spacing

Question:

Are APA events that shift toward distal sites associated with greater distances between APA sites compared to those shifting toward proximal sites?

What this tells you:

Whether distal shifts require greater transcript remodeling, such as longer 3′ UTRs or switching to far-away terminal exons.

If proximal shifts are more conservative, involving nearby APA choices.

statistically test whether APA events shifting toward distal sites differ in their inter-polyA distances compared to those shifting proximal, the Wilcoxon rank-sum test (Mann–Whitney U) is the right choice — especially since distances are often non-normally distributed.
```{r}
library(ggridges)

FR_sig_apa <- FR_sig_apa %>%
  mutate(APA_Direction = ifelse(deltapsi > 0, "Distal", "Proximal"),
         Population = "French")

ZA_sig_apa <- ZA_sig_apa %>%
  mutate(APA_Direction = ifelse(deltapsi > 0, "Distal", "Proximal"),
         Population = "Zambian")


apa_dist_data <- bind_rows(FR_sig_apa, ZA_sig_apa) %>%
  filter(numberofposfactors > 1, !is.na(interpolyAdist)) %>%
  mutate(interpolyAdist = as.numeric(interpolyAdist))

apa_dist_data_clean <- apa_dist_data %>%
  filter(!is.na(interpolyAdist), numberofposfactors > 1)


wilcox.test(interpolyAdist ~ APA_Direction, data = filter(apa_dist_data_clean, Population == "French"))
wilcox.test(interpolyAdist ~ APA_Direction, data = filter(apa_dist_data_clean, Population == "Zambian"))

wilcox.test(interpolyAdist ~ Population,
            data = filter(apa_dist_data_clean, APA_Direction == "Distal"))

wilcox.test(interpolyAdist ~ Population,
            data = filter(apa_dist_data_clean, APA_Direction == "Proximal"))



base_colors <- c("Proximal" = "pink", "Distal" = "darkblue")

ggplot(apa_dist_data, aes(x = interpolyAdist, y = interaction(Population, APA_Direction), fill = APA_Direction)) +
  geom_density_ridges(scale = 1.2, alpha = 0.8, rel_min_height = 0.01) +
  labs(title = "Distribution of APA distances",
       x = "Distance between APA sites (nt)",
       y = "APA Direction within Population") +
  theme_classic(base_size = 14) + 
  scale_color_manual(values = base_colors, labels = legend_labels) +
  theme(legend.position = "right", plot.title = element_text(face = "bold", hjust = 0.5))

```

```{r}
FR_sig_apa <- bind_rows(
  FR109_sig %>% mutate(Genotype = "FR109"),
  FR112_sig %>% mutate(Genotype = "FR112"),
  FR113_sig %>% mutate(Genotype = "FR113")) %>%
  filter(FDR < 0.05) %>%
  mutate(APA_Direction = ifelse(deltapsi > 0, "Distal", "Proximal"))

ZA_sig_apa <- bind_rows(
  ZI274_sig %>% mutate(Genotype = "ZI274"),
  ZI31_sig  %>% mutate(Genotype = "ZI31"),
  ZI418_sig %>% mutate(Genotype = "ZI418")) %>%
  filter(FDR < 0.05) %>%
  mutate(APA_Direction = ifelse(deltapsi > 0, "Distal", "Proximal"))


apa_dist_data <- bind_rows(FR_sig_apa, ZA_sig_apa) %>%
  filter(numberofposfactors > 1, !is.na(interpolyAdist)) %>%
  mutate(interpolyAdist = as.numeric(interpolyAdist))




base_colors <- c("Proximal" = "#fa8775", "Distal" = "#0000ff")

ggplot(apa_dist_data, aes(x = interpolyAdist, y = interaction(Genotype, APA_Direction), fill = APA_Direction)) +
  geom_density_ridges(scale = 1.2, alpha = 0.8, rel_min_height = 0.01, color = "black") +
  scale_fill_manual(values = base_colors) +
  labs(title = "Distribution of APA Distances by Genotype and Direction", x = "Distance between APA sites (nt)", y = "APA Direction per Genotype") +
  theme_classic(base_size = 14) +
  theme(legend.position = "right", plot.title = element_text(face = "bold", hjust = 0.5), axis.text.y = element_text(size = 10))

```



```{r}

# Extract validated IDs from both sets
fr_ids <- FR_sig_apa$validated_id
za_ids <- ZA_sig_apa$validated_id

# Genes private to French population
FR_private <- FR_sig_apa %>%
  filter(!(validated_id %in% za_ids)) %>%
  mutate(Population = "French")

# Genes private to Zambian population
ZA_private <- ZA_sig_apa %>%
  filter(!(validated_id %in% fr_ids)) %>%
  mutate(Population = "Zambian")

private_apa <- bind_rows(FR_private, ZA_private)

# View summary
table(private_apa$Population)



private_apa <- private_apa %>%
  mutate(ChangeType = case_when(
    deltapsi > 0 ~ "Distal",
    deltapsi < 0 ~ "Proximal",
    TRUE ~ NA_character_))
table(private_apa$Population, private_apa$ChangeType)


#Genetype Breakdown (TUTR, ALE, Mixed)

table(private_apa$Population, private_apa$genetype)


#APA Distance Summary (Mean, Median, SD by Population & ChangeType)

private_apa$interpolyAdist <- as.numeric(private_apa$interpolyAdist)


private_apa |>
  filter(!is.na(interpolyAdist)) |>
  group_by(Population, ChangeType) |>
  summarise(
    mean_distance = round(mean(interpolyAdist, na.rm = TRUE), 1),
    median_distance = round(median(interpolyAdist, na.rm = TRUE), 1),
    sd_distance = round(sd(interpolyAdist, na.rm = TRUE), 1),
    .groups = "drop")

private_apa %>%
  filter(genetype == "TUTR") %>%
  count(Population, ChangeType)

private_apa %>%
  filter(genetype == "ALE") %>%
  count(Population, ChangeType)


private_apa %>%
  filter(genetype == "mixed") %>%
  count(Population, ChangeType)
```


# genes that shared in the same direction
```{r}
FR_sig_apa <- FR_sig_apa %>%
  mutate(ChangeType = case_when(
    deltapsi > 0 ~ "Distal",
    deltapsi < 0 ~ "Proximal",
    TRUE ~ NA_character_))

ZA_sig_apa <- ZA_sig_apa %>%
  mutate(ChangeType = case_when(
    deltapsi > 0 ~ "Distal",
    deltapsi < 0 ~ "Proximal",
    TRUE ~ NA_character_))

shared_apa <- inner_join(
  FR_sig_apa %>% select(validated_id, current_symbol, ChangeType_FR = ChangeType),
  ZA_sig_apa %>% select(validated_id, current_symbol, ChangeType_ZA = ChangeType),
  by = "validated_id")

shared_same_direction <- shared_apa %>%
  filter(ChangeType_FR == ChangeType_ZA)

table(shared_same_direction$ChangeType_FR)
nrow(shared_same_direction)


shared_opposite_direction <- shared_apa %>%
  filter(ChangeType_FR != ChangeType_ZA)

shared_opposite_direction_unique <- shared_opposite_direction %>%
  distinct(validated_id, .keep_all = TRUE)

```

```{r}

private_apa <- private_apa %>%
  mutate(ChangeType = case_when(
    deltapsi > 0 ~ "Distal",
    deltapsi < 0 ~ "Proximal",
    TRUE ~ NA_character_))


# Genes unique to each population
FR_private <- FR_sig_apa %>%
  filter(!(validated_id %in% ZA_sig_apa$validated_id)) %>%
  mutate(Population = "French")

ZA_private <- ZA_sig_apa %>%
  filter(!(validated_id %in% FR_sig_apa$validated_id)) %>%
  mutate(Population = "Zambian")

# Combine
private_apa <- bind_rows(FR_private, ZA_private)

private_apa %>%
  filter(!is.na(interpolyAdist)) %>%
  group_by(Population, ChangeType) %>%
  summarise(
    mean_distance = round(mean(interpolyAdist, na.rm = TRUE), 1),
    median_distance = round(median(interpolyAdist, na.rm = TRUE), 1),
    sd_distance = round(sd(interpolyAdist, na.rm = TRUE), 1),
    .groups = "drop")

```



| Question                                                                    | What to Test or Compare |
| --------------------------------------------------------------------------- | --------------------- |
| **Are the distributions of APA site counts different between populations?** | Chi-squared or Fisher’s exact test on APA site frequencies   |
| **Do Zambian genes tend to have more APA sites?**                           | Compare mean/median APA site count (Wilcoxon test or t-test) |
| **What proportion of significant genes have exactly 2 APA sites?**          | Proportions, chi-squared test
| **Do APA site distances differ between populations (for 2-site genes)?**.   | Compare `interpolyAdist` (Wilcoxon test)|
| **APA complexity vs gene regulation direction (Δψ)** | Logistic regression, association analysis |



Interpretability of Δψ (delta psi) is most meaningful with two sites
For genes with two APA sites, the Δψ value clearly reflects a shift between the proximal site and the distal site.

A Δψ > 0 → increased usage of the proximal site (shortened UTR)
A Δψ < 0 → increased usage of the distal site (lengthened UTR)

With more than 2 sites, Δψ becomes a weighted average across multiple sites, and it's harder to biologically interpret the direction of the shift.



```{r}
private_apa <- bind_rows(
  FR109_sig,
  FR112_sig,
  FR113_sig,
  ZI274_sig,
  ZI31_sig,
  ZI418_sig)


dot_data <- private_apa %>%
  filter(!is.na(deltapsi), !is.na(current_symbol)) %>%
  mutate(
    ChangeType = case_when(
      FDR < 0.05 & deltapsi > 0 ~ "Proximal",
      FDR < 0.05 & deltapsi < 0 ~ "Distal",
      TRUE ~ "Not significant"),
    interpolyAdist = as.numeric(interpolyAdist) ) %>%
  arrange(desc(abs(deltapsi))) %>%
  distinct(current_symbol, .keep_all = TRUE) %>%
  slice_max(abs(deltapsi), n = 50) %>%
  mutate(current_symbol = factor(current_symbol, levels = current_symbol[order(deltapsi)]))


change_colors <- c("Proximal" = "pink", "Distal" = "darkblue")

ggplot(dot_data, aes(x = deltapsi, y = current_symbol, color = ChangeType, size = interpolyAdist, shape = Genotype)) + geom_point(alpha = 0.85) +
  scale_shape_manual(values = c("FR109" = 16, "FR112" = 17, "FR113" = 15, "ZI274" = 18, "ZI31" = 8, "ZI418" = 3)) + scale_size_continuous(range = c(2, 6), name = "APA Distance (nt)") +
  labs(title = "APA Changes in Significant Genes by Genotype", x = expression(Delta*psi~"(ψ[NR] - ψ[ER])"), y = "Gene", color = "Change Type",
    shape = "Genotype") +
  theme_classic(base_size = 14) +
  scale_color_manual(values = change_colors) +
  theme(axis.text.y = element_text(size = 7),plot.title = element_text(face = "bold", hjust = 0.5), legend.position = "right")

```




```{r}
# Keep only significant APA genes with genetype
FR109_genes <- FR109_sig %>% filter(!is.na(genetype)) %>%
  select(Gene, genetype) %>% mutate(Genotype = "FR109")

FR112_genes <- FR112_sig %>% filter(!is.na(genetype)) %>%
  select(Gene, genetype) %>% mutate(Genotype = "FR112")

FR113_genes <- FR113_sig %>% filter(!is.na(genetype)) %>%
  select(Gene, genetype) %>% mutate(Genotype = "FR113")

ZI274_genes <- ZI274_sig %>% filter(!is.na(genetype)) %>%
  select(Gene, genetype) %>% mutate(Genotype = "ZI274")

ZI31_genes <- ZI31_sig %>% filter(!is.na(genetype)) %>%
  select(Gene, genetype) %>% mutate(Genotype = "ZI31")

ZI418_genes <- ZI418_sig %>% filter(!is.na(genetype)) %>%
  select(Gene, genetype) %>% mutate(Genotype = "ZI418")



# Combine all genotypes
combined_genes <- bind_rows(
  FR109_genes, FR112_genes, FR113_genes,
  ZI274_genes, ZI31_genes, ZI418_genes)

# Count how many genotypes each gene is found in, per genetype
shared_counts <- combined_genes %>%
  distinct(Genotype, Gene, genetype) %>%
  group_by(Gene, genetype) %>%
  summarise(n_genotypes = n(), .groups = "drop")


# Count how many genes appear in N genotypes per genetype
genetype_sharing <- shared_counts %>%
  count(genetype, n_genotypes) %>%
  rename(`Number of Genotypes` = n_genotypes, `Number of Genes` = n)

```


```{r}
# Get vectors of significant gene IDs from each population
fr_ids <- unique(FR_collapsed$current_symbol)
za_ids <- unique(ZA_collapsed$current_symbol)
# Shared genes
shared_genes <- intersect(fr_ids, za_ids)

# French-only genes
french_only_genes <- setdiff(fr_ids, za_ids)

# Zambian-only genes
zambian_only_genes <- setdiff(za_ids, fr_ids)


private_apa %>%
  group_by(current_symbol) %>%
  tally() %>%
  dplyr::filter(n > 1)

```



LABRAT.py --conditionA Control --conditionB Ethanol
Then:

ΔΨ > 0 means ethanol-treated samples use the distal site more than control

ΔΨ < 0 means ethanol-treated samples shift to the proximal site


```{r, message=F, warning=F}
#FR109_filtered
#FR112_filtered 
#FR113_filtered
#ZI274_filtered
#ZI31_filtered
#ZI418_filtered

base_colors <- c("Proximal" = "pink", "Distal" = "darkblue")

FR109_box <- FR109_filtered %>%
  mutate(Population = "French_FR109",
         Direction = ifelse(deltapsi >= 0, "Distal", "Proximal"))

FR112_box <- FR112_filtered %>%
  mutate(Population = "French_FR112",
         Direction = ifelse(deltapsi >= 0, "Distal", "Proximal"))

FR113_box <- FR113_filtered %>%
  mutate(Population = "French_FR113",
         Direction = ifelse(deltapsi >= 0, "Distal", "Proximal"))


box_data <- bind_rows(FR109_box, FR112_box, FR113_box)


box_data$Direction <- factor(box_data$Direction, levels = c("Distal", "Proximal"))
box_data$Population <- factor(box_data$Population, levels = c("French_FR109", "French_FR112", "French_FR113"))




comp_list <- list(
  c("French_FR109", "French_FR112"),
  c("French_FR109", "French_FR113"),
  c("French_FR112", "French_FR113"))


ggplot(box_data, aes(x = Population, y = deltapsi, fill = Population)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = TRUE) +
  geom_jitter(width = 0.2, alpha = 0.6, color = "black", size = 1) +
  stat_compare_means(comparisons = comp_list, method = "wilcox.test", label = "p.signif", aes(group = Population)) +
  facet_wrap(~Direction, ncol = 1, scales = "free_y", strip.position = "right") +
  labs(title = "APA shift across French Replicates", y = expression(Delta*psi~"(ψ[NR] - ψ[ER])"), x = "") +
  scale_color_manual(values = base_colors, labels = legend_labels) + ylim(-0.3, 0.3) +
  theme_classic(base_size = 14) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"), plot.title = element_text(face = "bold", hjust = 0.5)) +
  coord_flip()


```


```{r, message=F, warning=F}

box_data$Direction <- factor(box_data$Direction, levels = c("Distal", "Proximal"))
box_data$Population <- factor(box_data$Population)

direction_comparison <- list(c("Distal", "Proximal"))


ggplot(box_data, aes(x = Direction, y = deltapsi, fill = Direction)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = TRUE) +
  geom_jitter(width = 0.2, alpha = 0.7, color = "black", size = 1) +
  stat_compare_means(comparisons = direction_comparison, method = "wilcox.test", label = "p.signif", label.y = 0.1) +
  facet_wrap(~Population, ncol = 1, strip.position = "right") +
  labs(title = "Δψ within French replicates", y = expression(Delta*psi~"(ψ[NR] - ψ[ER])"), x = "APA Direction") +
  scale_fill_manual(values = base_colors, labels = legend_labels) + ylim(-0.3, 0.3) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  coord_flip()

```


```{r, message=F, warning=F}
ZI274_box <- ZI274_filtered %>%
  mutate(Population = "Zambian_ZI274",
         Direction = ifelse(deltapsi >= 0, "Distal", "Proximal"))

ZI31_box <- ZI31_filtered %>%
  mutate(Population = "Zambian_ZI31",
         Direction = ifelse(deltapsi >= 0, "Distal", "Proximal"))

ZI418_box <- ZI418_filtered %>%
  mutate(Population = "Zambian_ZI418",
         Direction = ifelse(deltapsi >= 0, "Distal", "Proximal"))


box_data <- bind_rows(ZI274_box, ZI31_box, ZI418_box)


box_data$Direction <- factor(box_data$Direction, levels = c("Distal", "Proximal"))
box_data$Population <- factor(box_data$Population, levels = c("Zambian_ZI274", "Zambian_ZI31", "Zambian_ZI418"))

# Define all pairwise comparisons you want
comp_list <- list(
  c("Zambian_ZI274", "Zambian_ZI31"),
  c("Zambian_ZI274", "Zambian_ZI418"),
  c("Zambian_ZI31", "Zambian_ZI418"))


ggplot(box_data, aes(x = Population, y = deltapsi, fill = Population)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = TRUE) +
  geom_jitter(width = 0.2, alpha = 0.6, color = "black", size = 1) +
  stat_compare_means(comparisons = comp_list, method = "wilcox.test", 
                     label = "p.signif", aes(group = Population)) +
  facet_wrap(~Direction, ncol = 1, scales = "free_y", strip.position = "right") +
  labs(title = "APA shift across Zambian replicates",
       y = expression(Delta*psi~"(ψ[NR] - ψ[ER])"), x = "") +
  scale_color_manual(values = base_colors)  +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  coord_flip()
```


```{r, message=F, warning=F}
box_data$Direction <- factor(box_data$Direction, levels = c("Distal", "Proximal"))
box_data$Population <- factor(box_data$Population)

direction_comparison <- list(c("Distal", "Proximal"))

#Within_French_Psi
ggplot(box_data, aes(x = Direction, y = deltapsi, fill = Direction)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6, notch = TRUE) +
  geom_jitter(width = 0.2, alpha = 0.7, color = "black", size = 1) +
  stat_compare_means(comparisons = direction_comparison, method = "wilcox.test", label = "p.signif", label.y = 0.1) +
  facet_wrap(~Population, ncol = 1, strip.position = "right") +
  labs(title = "Δψ within Zambian replicates", y = expression(Delta*psi~"(ψ[NR] - ψ[ER])"), x = "APA Direction") +
 scale_fill_manual(values = base_colors) + ylim(-0.3, 0.3) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)) +
  coord_flip()

```


Identify genes that show both APA and A3SS regulation, for each population.
```{r}
A3SS_Zambian= read.delim("~/Dropbox/3UTR/A3SS_Zambian.txt", header = T, sep = "\t", check.names = T)
A3SS_French = read.delim("~/Dropbox/3UTR/A3SS_French.txt", header = T, sep = "\t", check.names = T)


A3SS_Zambian_filtered = A3SS_Zambian |> 
   filter(FDR<= 0.05)

A3SS_French_filtered = A3SS_French |> 
  filter(FDR <= 0.05)


```

```{r}
all_sig_genes_FR <- bind_rows(FR109_filtered, FR112_filtered,FR113_filtered)

all_sig_genes_ZA <- bind_rows(ZI274_filtered, ZI31_filtered,ZI418_filtered)



all_sig_genes_FR = all_sig_genes_FR %>%
  distinct(validated_id, .keep_all = T)

all_sig_genes_ZA = all_sig_genes_ZA %>%
  distinct(validated_id, .keep_all = T)


a3ss_za_ids <- unique(A3SS_Zambian$GeneID) 
a3ss_fr_ids <- unique(A3SS_French$GeneID)
apa_za_ids <- unique(all_sig_genes_ZA$Gene)
apa_fr_ids <- unique(all_sig_genes_FR$validated_id)




# Overlap between APA and A3SS genes
overlap_za <- intersect(a3ss_za_ids, apa_za_ids)
overlap_fr <- intersect(a3ss_fr_ids, apa_fr_ids)

# Counts
length(overlap_za)
length(overlap_fr)

```


```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

